Generalized Plackett-Luce Likelihoods

### 2022-08-14 inst/jss_hyper3.Rnw
minor tidyup

<<>>=
H <- hyper3()
H[c(a=1)] <- 1
H[c(a=3,b=2,c=1)] <- -1
H
@

Above, we see object \code{H} is a log-likelihood function of the
players' strengths, which may be evaluated at specified points in
probability space.  If, for example, we wish to assess $H_1\colon
p_a=0.9,p_b=0.05,p_c=0.05$ and $H_2\colon p_a=0.01,p_b=0.01,p_c=0.98$,
we can evaluate these hypotheses using generic function
\code{loglik()}:

<<>>=
loglik(c(a=0.01,b=0.01,c=0.98),H)
loglik(c(a=0.9,b=0.05,c=0.05),H)
@

@@

We may use a selection of statistical tests on the order statistic,
modified from~\citep{hankin2017} to work with \code{hyper3} objects.
For example, we may assess the hypothesis that all three players are
of equal strength [viz $H_0\colon p_a=p_b=p_c=\frac{1}{3}$]:

@@

Here I consider some order statistics with nontrivial maximum
likelihood Bradley-Terry strength.  The simplest nontrivial case would
be three competitors with strengths $a,a,b$ and finishing order
$a\succ b\succ a$.  The likelihood function would be

@@

This will be zero at $\sqrt{2}\pm 1$; we also note that
$d^2\mathcal{L}/da^2=-4(1+a)^{-3}$, manifestly strictly negative for
$0\leq a\leq 1$: the root is a maximum.

@@

above, we see close agreement with the theoretical value of
$(\sqrt{2}-1,2-\sqrt{2})\simeq (0.414,0.586)$.  Observe that the
maximum likelihood estimate for $a$ is strictly less than 0.5, even
though the finishing order is symmetric.  Using
$\mathcal{L}(a)=\frac{a(1-a)}{1+a}$, we can show that
$\log\mathcal{L}(\hat{a})=\log\left(3-2\sqrt{2}\right)\simeq -1.76$,
where $\hat{a}=\sqrt{2}-1$ is the maximum likelihood estimate for $a$.
Defining $\mathcal{S}=\log\mathcal{L}$ as the support [log-likelihood]
we have

@@

The $p$-value is about 0.81, exceeding 0.05; thus we have no strong
evidence to reject the null of $a=\frac{1}{2}$.  The observation is
informative, in the sense that we can find a credible interval for
$a$.  With an $n$-unit of support criterion the analytical solution to
$\mathcal{S}(p)=-n$ is given by defining $X=\log(3-2\sqrt{2})-n$ and
solving $p(1-p)/(1+p)=X$, or
$p_\pm=\left(1-X\pm\sqrt{1+4X+X^2}\right)/2$, the two roots being the
lower and upper limits of the credible interval;
Figure~\ref{ABA_likelihood}.

@@

The ideas presented above can easily be extended to arbitrarily large
numbers of competitors, although the analytical expressions tend to be
intractable and numerical methods must be used.  All results and
datasets presented here are maintained under version control and
available at \url{https:github.com/RobinHankin/hyper2}.  Given an
order statistic of the type considered above, the
Mann-Whitney-Wilcoxon test~\citep{mann1947,wilcoxon1945} assesses a
null of identity of underlying distributions.  Consider the
chorioamnion dataset at \code{wilcox.test.Rd}:

@@

Here we see permeability constants of the human chorioamnion at term
(\code{x}) and between 12 to 26 weeks gestational age (\code{y}).  The
order statistic is straightforward to calculate:

@@

A particularly attractive feature of this analysis is that the
Bradley-Terry strengths have direct operational significance: If two
competitors, say Vadlejch and Vesely, were to throw a javelin, then we
would estimate the probability that Vadlejch would throw further than
Vesely at $\displaystyle p_{\mbox{\tiny Vad}}/\left(p_{\mbox{\tiny
Vad}} + p_{\mbox{\tiny Ves}}\right)\simeq 0.74$.  Indeed, from a
training or selection perspective we might follow \cite{hankin2017}
and observe that log-contrasts~\citep{ohagan2004}
$\log\left(p_{\mbox{\tiny Vad}}/ p_{\mbox{\tiny Ves}}\right)$ appear
to have approximately Gaussian likelihood functions for observations
of the type considered here.  Profile log-likelihood curves for
log-contrasts are easily produced by the package,
Figure~\ref{profliklogcont}.

@@

In Formula 1 motor racing, the constructors' championship takes place
between {\em manufacturers} of racing cars (compare the drivers'
championship, which is between drivers).  In this analysis, the
constructor is the object of inference.  Each constructor typically
fields two cars, each of which separately accumulates ranking-based
points at each venue.  Here we use generalized Plackett-Luce to assess
the contructors' performance.  The following table, included in the
\code{hyper2} package as a dataset, shows rankings for the first 9
venues of the 2021 season:

@@

New likelihood functions for simple cases with three or four
competitors were presented, and extending to larger numbers furnishes
a generalization of the Mann-Whitney-Wilcoxon test that offers a
specific alternative (Bradley-Terry strength) with a clear operational
definition.  Further generalizations allow the analysis of more than
two groups, here applied to Olympic Javelin throw distances.
Generalized Plackett-Luce likelihood functions were used to assess the
constructors' championship and a reasonable null.  Specifically, the
hypothesis that the strength of the Mercedes team remained unchanged
between 2020 and 2021 was tested and rejected.

>>>>

<<>>=
LL <- hyper3()
LL[c(a=1)] <- 1
LL[c(a=3,b=2,c=1)] <- -1
LL
@

Above, we see object \code{LL} is a log-likelihood function of the
players' strengths, which may be evaluated at specified points in
probability space.  If, for example, we wish to assess $H_1\colon
p_a=0.9,p_b=0.05,p_c=0.05$ and $H_2\colon p_a=0.01,p_b=0.01,p_c=0.98$,
we can evaluate these hypotheses using generic function
\code{loglik()}:

<<>>=
loglik(c(a=0.01,b=0.01,c=0.98),LL)
loglik(c(a=0.9,b=0.05,c=0.05),LL)
@

@@

The package provides a number of statistical tests on likelihood
functions, modified from~\citep{hankin2017} to work with \code{hyper3}
objects.  For example, we may assess the hypothesis that all three
players are of equal strength [viz $H_0\colon
p_a=p_b=p_c=\frac{1}{3}$]:

@@

Here I consider some order statistics with nontrivial maximum
likelihood Bradley-Terry strengths.  The simplest nontrivial case
would be three competitors with strengths $a,a,b$ and finishing order
$a\succ b\succ a$.  The likelihood function would be

@@

This will be zero at $\sqrt{2}-1$; we also note that
$d^2\mathcal{L}/da^2=-4(1+a)^{-3}$, manifestly strictly negative for
$0\leq a\leq 1$: the root is a maximum.

@@

Above, we see close agreement with the theoretical value of
$(\sqrt{2}-1,2-\sqrt{2})\simeq (0.414,0.586)$.  Observe that the
maximum likelihood estimate for $a$ is strictly less than 0.5, even
though the finishing order is symmetric.  Using
$\mathcal{L}(a)=\frac{a(1-a)}{1+a}$, we can show that
$\log\mathcal{L}(\hat{a})=\log\left(3-2\sqrt{2}\right)\simeq -1.76$,
where $\hat{a}=\sqrt{2}-1$ is the maximum likelihood estimate for $a$.
Defining $\mathcal{S}=\log\mathcal{L}$ as the support [log-likelihood]
we have

@@

The $p$-value is about 0.81, exceeding 0.05; thus we have no strong
evidence to reject the null of $a=\frac{1}{2}$.  The observation is
informative, in the sense that we can find a credible interval for
$a$.  With an $n$-units of support criterion the analytical solution to
$\mathcal{S}(p)=-n$ is given by defining $X=\log(3-2\sqrt{2})-n$ and
solving $p(1-p)/(1+p)=X$, or
$p_\pm=\left(1-X\pm\sqrt{1+4X+X^2}\right)/2$, the two roots being the
lower and upper limits of the credible interval;
Figure~\ref{ABA_likelihood}.

@@

The ideas presented above can easily be extended to arbitrarily large
numbers of competitors, although the analytical expressions tend to be
intractable and numerical methods must be used.  All results and
datasets presented here are maintained under version control and
available at \url{https:github.com/RobinHankin/hyper2}.  Given an
order statistic of the type considered above, the
Mann-Whitney-Wilcoxon test~\citep{mann1947,wilcoxon1945} assesses a
null of identity of underlying distributions.  Consider the
chorioamnion dataset~\citep{hollander2013}, used in
\code{wilcox.test.Rd}:

@@

Here we see a measure of permeability of the human placenta at term
(\code{x}) and between 3 and 6 months' gestational age (\code{y}).
The order statistic is straightforward to calculate:

@@

A particularly attractive feature of this analysis is that the
Bradley-Terry strengths have direct operational significance: If two
competitors, say Vadlejch and Vesely, were to throw a javelin, then we
would estimate the probability that Vadlejch would throw further than
Vesely at $\displaystyle p_{\mbox{\tiny Vad}}/\left(p_{\mbox{\tiny
Vad}} + p_{\mbox{\tiny Ves}}\right)\simeq 0.74$.  Indeed, from a
training or selection perspective we might follow \cite{hankin2017}
and observe that log-contrasts~\citep{ohagan2004} appear to have
approximately Gaussian likelihood functions for observations of the
type considered here.  Profile log-likelihood curves for log-contrasts
are easily produced by the package, Figure~\ref{profliklogcont}.  We
see that the credible range for~$\log\left(p_{\mbox{\tiny Vad}}/
p_{\mbox{\tiny Ves}}\right)$ includes zero and we have no strong
evidence for these athletes having different (Bradley-Terry)
strengths.

@@

Formula 1 motor racing is an important and prestigious motor sport
\citep{codling2017,jenkins2010}.  In Formula 1 Grand Prix, the
constructors' championship takes place between {\em manufacturers} of
racing cars (compare the drivers' championship, which is between
drivers).  In this analysis, the constructor is the object of
inference.  Each constructor typically fields two cars, each of which
separately accumulates ranking-based points at each venue.  Here we
use generalized Plackett-Luce to assess the contructors' performance.
The following table, included in the \code{hyper2} package as a
dataset, shows rankings for the first 9 venues of the 2021 season:

@@

New likelihood functions for simple cases with three or four
competitors were presented, and extending to larger numbers furnishes
a generalization of the Mann-Whitney-Wilcoxon test that offers a
specific alternative (Bradley-Terry strength) with a clear operational
definition.  Further generalizations allow the analysis of more than
two groups, here applied to Olympic javelin throw distances.
Generalized Plackett-Luce likelihood functions were used to assess the
Grand Prix constructors' championship and a reasonable null.
Specifically, the hypothesis that the strength of the Mercedes team
remained unchanged between 2020 and 2021 was tested and rejected.
